---
title: "Example Markdown"
author: "tim"
date: "11/11/2019"
output: pdf_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introducing `R markdown`

## This is a next level subsection

The basic idea with markodown is that your code stays with the text that it pertains to. Wow. That can mean different things. 

1. It could just mean that your research scripts are documented. Meaning, the code that you write will not fall into a state of mystery.

Here is a little chunk making ten random normal draws.
```{r}
a <- rnorm(10)
```

Now we can do different things with `a`.

```{r, echo = FALSE, results = "hide", message = FALSE, warning=FALSE}
a^2
plot(a)
```

Here we talked about different chunk options:

1. `echo = FALSE` hide code
2. `results = "hide"` hide console
3. `message = FALSE` supppress messages
4. `warning = FALSE` suppress warnings
5. `eval = FALSE` don't execute the chunk

# Tidy concepts

Tidy data is formatted to have observations in rows and variables in columns. The meaning of observation depends on the downstream analysis. It might be years (aggregates), or years and ages, or individuals, or individual-waves, or other things. So depending on the task at hand, the tidy format changes.

So, if our task is get data from the source to a tidy format, we need to be clear about the objectives.

# An example of tidy data: `gapminder`

```{r, results = "hide", warning=FALSE,message=FALSE}
# install.packages("gapminder")
# install.packages("tidyverse")
# install.packages("here")
library(gapminder)
library(tidyverse)
library(here)
dim(gapminder)
head(gapminder)
```

The following intro is a paraphrased and distilled version of [@healy2018data]. This book has both a package and a free website [http://socviz.co/](http://socviz.co/), so you don't need to buy it. 

Gapminder is tidy in the sense that it's designed to make Preston curves (LE by GDPpercap scatterplots), an observation is defined as a country-year combination. It's a nice dataset to showcase basic `ggplot2`.

```{r}
head(gapminder)
ggplot(gapminder, 
       mapping = aes(x = gdpPercap,
                     y = lifeExp,
                     color = continent)) +
  geom_point()
```

The important things: first feed it a tidy dataset, then declare a `mapping` of variables to visual attributes, which first and foremost define a coordinate mapping and optionally also other visual aesthetics that need mapping, like assigning qualitative colors to the continents. In this case `ggplot()` detects that the continent variable is character and therefore defaults to a qualitative palette.

## line plot panel by continent.

```{r}
head(gapminder)
library(gghighlight)
#install.packages("gghighlight")
ggplot(gapminder,
       mapping = aes(x = year,
                     y = lifeExp,
                     group = country,
                     color = country)) + 
  geom_line() +
  gghighlight(max(lifeExp) > 82) +
  geom_line(filter(gapminder, country == "China"),
            mapping = aes(x = year,
                     y = lifeExp),
            size = 1.5,
            color = "red") + 
  annotate(geom = "text",
           x = 1970,
           y = 58,
           label = "China")
```

Here we see how to use `gghighlight` to highlight a few countries that meet a condition atop all other series, which are identically rendered in the *background*. In the first iteration we had a bunch of black lines facetted. I brought up how people often foolishly try to have legends for colored lines in such situations. Don't do that. Instead make the non-relevent ones fall to the background and highlight and bold the series that you actually discuss in the text. We do this in two ways: 1) `geom_highlight()`, 2) by subsetting `gapminder` and asssigning a new mapping in a second geom. Lesson: geoms can have their own mappings. This is a common thing to trip up on.

# `dplyr` a worked example

Let's download this spreadsheet and figure out how to work with it!

[http://data.un.org/DocumentData.aspx?id=319](http://data.un.org/DocumentData.aspx?id=319)

Download the spreadsheet and stick it in a new folder called `Data`. We're going to figure this thing out.

```{r}
library(readxl)
library(readxl)
ASFR <- read_excel(
  "Data/Age-specific fertility rates, Total fertility and .xls", 
    na = "..", 
    skip = 4) %>% # Ctrl + Shift + m
  select(1:13) %>% 
  dplyr::rename("Country" = "...1", "ISO" = "...2",
                     "PeriodVerbose" = "...3","TFR" = "...6",
                     "15" = "15-19","20" = "20-24",
                     "25" = "25-29","30" = "30-34",
                     "35" = "35-39","40" = "40-44",
                     "45" = "45-49")
```

Here we have a 3-step pipeline! Yay. The first step is syntax to read in from a spreadsheet, generated interactively using the import data feature of rstudio. Copy. Paste. Bam. Step 2 `select()`s just the columns we want for now. Step 3 `rename()`s the columns manually to something useful. The pipe operator `%>%` take the object coming out of the function on the left and feeds it to the function on the right (by default as its first argument). These steps can be strung together into so-called data processing *pipelines*. 

Pipelines read like sentences. Because they are made of verbs and the pipe operator reads as *and then do this*.

```{r}
# here we're assigning at the top, but it's getting
# what comes out the bottom
ASFR <- ASFR %>% 
  select(-c(PeriodVerbose, TFR, Period)) %>% 
  pivot_longer(cols = `15`:`45`, 
               names_to = "Age",
               values_to = "asfr") %>% 
  mutate(Age = as.integer(Age),
         asfr = asfr / 1000) %>% 
  group_by(Country, Year) %>% 
  mutate(TFR = sum(asfr, na.rm = TRUE) * 5,
         MAB = sum((5 * asfr) * (Age + 2.5), na.rm = TRUE) / TFR) %>%
  ungroup() %>% 
  filter(TFR > 0)  # ASFR up top gets what comes out here
```

The very first step could be written like this:
```{r, eval = FALSE}
  select(ASFR, -c(PeriodVerbose, TFR, Period))
```

What method do we use to split these data to single ages?

Suggestions from the field:

1. PCLM
2. Sprague (or Beers, or Grabill, or ...)
3. smooth spline through uniform repeated values in age groups (not constrained, but could work OK)
4. monotonic spline through cumulative distribution
5. use a fertility model, e.g. Coale-Trussel and pals
6. Schmertmann [http://schmert.net/calibrated-spline/](http://schmert.net/calibrated-spline/) Super awesome love it

We will try the monotonic spline method because it seems easier to write fresh.

```{r}
mono_graduate <- function(Age5, asfr5){
  
  # assume NAs are 0s, sometimes these are found 
  # in the first or last age groups.
  asfr5[is.na(asfr5)] <- 0
  # anchor the endpoints (added double anchor)
  Age5                <- c(13,14, Age5+4, 50,51)
  asfr5               <- c(0,0, asfr5, 0,0)
  # scale up asfr
  asfr5               <- asfr5 * 5
  # accumualte it
  casfr5              <- cumsum(asfr5)
  # start at 14 because we need firs differences
  # to accumulate...
  predict_ages        <- 14:49
  # fit the spline
  xy <- splinefun(x = Age5,
            y = casfr5,
            method = "monoH.FC")(predict_ages)
  # return tibble
  tibble(Age = 15:49, 
         asfr = diff(xy))
}
chunk <- filter(ASFR, ISO == 4, Year == 1973)
plot(mono_graduate(Age5 = chunk$Age,
              asfr5 = chunk$asfr))
```

Great now we have a spline function, let's see if we can figure out how to use it in a pipeline.

```{r, message=FALSE}
ASFR <- ASFR %>% 
 	group_by(Country, ISO, Year) %>% 
    do(data = mono_graduate(Age5 = .$Age, asfr5 = .$asfr)) %>% 
 	unnest(cols = c(data)) %>% 
  group_by(Country, Year) %>%
  mutate(TFR = sum(asfr),
         MAB = sum(asfr * (Age+.5))/TFR)

nrow(ASFR) # yup it's bigger now
```

Great, the trick is to be able to return data chunks in a new kind of column called a list column. Having a list column means you have nested data, however, we can get it back to flat tidy format by `unnest()`ing. Note that we lost TFR and MAB in this operation. A cheap way to keep them would have been to incldue them in the first `group_by()` statement, as I did for `ISO` (`ISO` and `Country` are redundant). Each subset now has single ages 15-49. We don't know yet whether there were any pathological cases. Let's make a scatterplot of MAB by TFR to find out.

```{r}
#install.packages("ggpointdensity")
library(ggpointdensity)
library(colorspace)
ASFR %>% 
  filter(Age == 30,
         MAB > 20) %>% 
  ggplot(mapping = aes(x = MAB, y = TFR)) +
  geom_pointdensity() + 
  scale_color_continuous_sequential(palette = "Reds")
```

Now we start with the exercise. We bin TFRs into some intervals, let's just say .25 or so. This can happen in a `mutate()` call, and we do it by subtracting the modulo using `%%` (super awesome operator). Next step is to group by `TFRint` and pick out the minumum and maximum MAB in each interval band.

```{r}

ASFRex <- ASFR %>% 
  filter(MAB > 20) %>% 
  mutate(TFRint = TFR - TFR %% .25) %>% 
  group_by(TFRint) %>% 
  mutate(extremes = case_when(
                      MAB == max(MAB) ~ "max",
                      MAB == min(MAB) ~ "min",
                      TRUE ~ as.character(NA)
  )) %>% 
 filter(!is.na(extremes))
```

Now this is something we can plot! Let's use the `ggridges` package to bring in the ridgeplot geom.

```{r, fig.height=15}
#install.packages("ggridges")
library(ggridges)

ASFRex %>% 
  filter(TFRint >= 1,
         TFRint < 7.1) %>% 
  ggplot(mapping = aes(x = Age, 
                y = factor(TFRint, 
 						 		           levels = sort(unique(TFRint)), 
 						 		           ordered = TRUE))) + 
 	geom_ridgeline(mapping = aes(x = Age, 
 								               height = asfr, 
 								               fill = extremes),  
 	               scale = 4, 
 				         alpha = .6) +
  labs(y = "TFR")

```

We start by filtering out some of the extreme TFRs, which have few observations. Then we pass the main data object into `ggplot()`, mapping `Age` to `x` and `y` to `TFRint`. In this case, `y` means the ridge level, not the shifted height of each curve. Note since we want TFR to increase on the `y` axis, we need to set it as an ordered factor. This is a common ggplot2 trick, so not bad to see it. To draw the curves, we use the `geom_ridgeline()` geom, giving a new mapping there. Within the geom, `x` is `Age`, and `y` is `asfr`, and we finally map fill color to the extremes variable. 

Looking at the results, we note a few odd cases, which could be problematic spline fits, or poor data quality. If we wanted to pursue this any further then these would require closer examination. I'm afraid some of these series might have poor quality, or else data gaps that cause the splines to behave bad, not sure. We could also try the Schmertmann or some other method. Let's chalk this up as a win for now! 

# References





















