---
pagetitle: "Combining Data Sources"
title: | 
  | Data Wrangling for EDSDers
  | \vspace{1.5cm} \LARGE\emph{Module outline}
author: |
  | United Nations Population Division
  | 11-14 Nov, 2019
  | Tim Riffe
  | Max-Planck-Institute for Demographic Research
date: "13 August, 2019"
output:
  html_document:
    number_sections: yes
    toc: yes
params:
  output_dir: "../EDSD2019data/docs"
header-includes:
- \usepackage{titling}
- \pretitle{\begin{center}\includegraphics[trim=0 0 0 8cm, width=6cm, ]{assets/MPIDR_square_color.pdf}\\[\bigskipamount]}
- \posttitle{\end{center}}
bibliography: references.bib
---


<a href="https://github.com/timriffe/EDSD2019data" class="github-corner" aria-label="View source on GitHub"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#70B7FD; color:#fff; position: absolute; top: 0; border: 0; right: 0;" aria-hidden="true"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

I've download the World Values Survey Longitudinal file from here: [http://www.worldvaluessurvey.org/](http://www.worldvaluessurvey.org/), which ships in 4 different data formats, including `.rds`! Download it too, and unzip it in the `Data` folder for this project. Let's see what's inside.

Packages we'll use:
```{r, message = FALSE, warning = FALSE}
library(here)
library(tidyverse)
```

This dataset is quite wide because they ask tons of questions. Many of them are super intersting. The objective now is to merge some variable from WVS into the a WPP demographic dataset. We need to solve the following problems in order to have merge variables that are sufficiently harmonized for this to work. Meaning: 

1. Age should either be in clean single ages or clean grouped ages. For similicity we'll use 5-year age groups. 
2. We need harmonized years. WPP give 5-year intervals, so we can year-group the WVS data.
3. We need to get some sort of prevalence or mean measure per 5-year age group in some interesting variable (by year, age, sex and country). We can do this using the `summarize()` function instead of regression. Regression would be cool though.
4. We need to reshape the WPP data to long (`pivot_longer`)
```{r}
WVS <- readRDS(
  here::here("Data",
             "F00008390-WVS_Longitudinal_1981_2016_r_v20180912.rds"))
colnames(WVS)
# country code
unique(WVS$S003)
# year
unique(WVS$S020)
# homogeneous weight (all countries equal N = 1500)
# S019 

# sex
unique(WVS$X001)

# age
hist(WVS$X003)
# cohort
WVS$X002

# A015 "People can be trusted"
table(WVS$A165)
```

Select ISO, Year, Age, Sex, Person Weight, Trust
```{r}
WVS %>% 
  select(ISO = S003, 
         year = S020, # Year
         sex = X001, # Sex
         age = X003, # age
         trust = A165,
         pwt = S017) %>%  # trust
  #pull(age) %>% table() # spot check for negative age
  filter(age >= 0) %>% 
  mutate(year = year - year %% 5,
         age = age - age %% 5) %>% 
  glimpse()
```




Now let's talk about what one might join and how. Here's a dataset that uses the same country codes.
```{r}
#install.packages("wpp2019")
library(wpp2019)
help(wpp2019)
#head(mxF)
# not bad coincidence!
intersect(unique(mxF$country_code),unique(WVS$S003))
```

